\plaintitle

### 2025-05-15 docs/style/article-body.tex
edited method

Further details about how this signal is produced and sampled will be introduced in section 2 and 3, and builds on the work of Clason \cite{Clason2023}. This optical signal needs to be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). Some processing of the data has to be done in order to ensure that the data is workable. As such, these random bits will be processed via Toeplitz extraction \cite{toeplitz} in order to process these bits into random numbers. These random numbers will then be output from the microcontroller to the host computer via USB. This thesis will aim to answer one key research question: How can sampled vacuum fluctuations be processed efficiently in order to output QRNG?

In producing this firmware, several key considerations have to be made in order for this system to be usable in a production environment. The vision for the end product is a simple USB-stick that can be connected to a host, and produce true random numbers from ambient quantum fluctuations. While this system could very well be implemented on physically larger hardware, thus avoiding the constraints that limited hardware introduces, Clason proposes a simpler and cheaper way to achieve QRNG \cite{Clason2023}. In keeping with this, utilizing microcontrollers for processing the raw bits extracted further helps to keep the costs low and the solution reasonably complex. Further details regarding the microcontrollers used in our work will be outlined in section 3. However, due to this portability constraint, our implementation needs to work quickly and efficiently on resource constrained hardware. As such, our main question is broken down into two concrete research areas:

@@

\textbf{Research area 2 (\emph{RA2})}: Can we ensure that the output of random numbers is not limited primarily by firmware, but rather limited by the processing power or the USB transfer speed of the microcontroller, alternatively by the ADC?

@@

Section 2 of this article will introduce the theory that allows for QRNG, and how this will be utilized in our works. Section 3 delves further into the hardware and algorithms our work will use, with related works in optimizing Toeplitz extraction listed under section 4. Section 5 will present our methodology as well as some limitations imposed on our work. \textbf{More sections to follow as we finish the article}.

@@

Microcontrollers (MCUs) are compact and low-power computing devices designed primary for embedded systems and real-time applications. Unlike general purpose CPUs, MCUs integrates a processor, memory, and peripherals into a single chip. Thus, enabling efficient and autonomous operation in constrained environments. Various development frameworks and tools are available for programming MCUs, including the Arduino Framework, manufacturer specific SDKs and raw C++, among others. Each framework presents different trade-offs in terms of performance overhead, efficiency, and ease of development.

Modern microcontrollers feature advanced microarchitectural elements to enhance processing capabilities. Some of high performance MCUs utilize dual-issue superscalar architectures, enabling them to run several instructions in one cycle. This type of design greatly improves single-threaded capabilities, which makes these MCUs ideal for tasks that need real-time data processing. The use of dual-issue architectures in embedded devices has demonstrated high execution speeds, along with the added benefit of improved energy efficiency. Thus, making them a great option for applications that require low power but high performance. Research conducted by Lozano et al. \cite{dual-issue} specifically explores energy efficient dual-issue processors designs. This study shows that optimizing instruction throughput and power consumption can significantly improve processing efficiency in embedded systems. While energy efficiency is not the primary goal for this study, ensuring that the resulting USB-device does not consume too much power is tangentially relevant for real-world usage.

@@

The Teensy 4.1\footnote{\href{https://www.pjrc.com/store/teensy41.html}{Teensy developer documentation, accessed 2025-02-27}}, based on the ARM Cortex-M7 is especially good for computationally demanding tasks involving randomness extraction due to its dual-issue superscalar architecture and DSP capabilities. The floating-point unit\footnote{\href{https://www.sciencedirect.com/topics/computer-science/floating-point-unit}{ScienceDirect Journals \& Books, accessed 2025-03-13}} (FPU) and SIMD-style DSP instructions improves how quickly it can perform bitwise and arithmetic tasks, which are crucial for quick Toeplitz extraction. SIMD-controlled DSP architectures, as described by Han et al. \cite{simd-dsp}, leverage parallel vectorized computation to accelerate matrix operations. Thus, making them highly effective for Toeplitz matrix-vector multiplications. The Cortex-M7 does not have special features for cryptography, its ability to execute multiple instructions quickly, combined with fast memory access and an efficient pipeline design, allows it to effectively manage and process randomness extraction. Moreover, its USB High-Speed (480 Mbit/sec) interface, supporting both USB device and USB host modes, allows rapid transmission of extracted random data to an external system, ensuring minimal bottlenecks in high-rate randomness generation.

In order to evaluate just how much constraint our implementation can allow, our aim is to try our implementation on other MCUs with varying levels of power and hardware support. Whereas Teensy 4.1 is our primary development platform which we will evaluate closely, we aim to run our implementations on Raspberry Pi Pico 2\footnote{\href{https://datasheets.raspberrypi.com/pico/pico-2-product-brief.pdf}{Raspberry Pi Pico 2 documentation, accessed 2025-03-13}} as well as ESP32-S3\footnote{\href{https://www.espressif.com/sites/default/files/documentation/esp32-s3_datasheet_en.pdf}{ESP32-S3 documentation, accessed 2025-03-13}}. Due to the lower computational power of these MCUs, there may be significant issues in utilizing these weaker models, yet they are significantly cheaper and easier to access. Testing of these will consist solely of running the implementation on these controllers and measuring execution speed.

\subsubsection{\texorpdfstring{3.4 Toeplitz extraction }{3.4 Toeplitz extraction }}\label{toeplitz-extraction}

@@

Let X be the input array (length n)
Let T be the seed array (length n + m – 1)
Let Y be a new array of length m

@@

Whereas this example is quite small in order to demonstrate how the algorithm operates, this can of course be scaled up significantly with larger values for both \(T\) and \(k\) respectively, granting larger integers. The main focus of this work is implementing this algorithm as efficiently as possible on our MCUs, and as such, several optimization efforts need to be taken into account during our experimentation.

@@

\subsection{\texorpdfstring{4 RELATED WORKS }{4 RELATED WORKS }}\label{related-works}

@@

However, in order to ensure the results work with varying data, we cannot limit ourselves to simply one stream of bits. The main point of the algorithm is to remove patterns in the bitstream that may lead to less randomized results. As such, we will manually produce several bitstreams to use for our tests -- each with varying degrees of repeated patterns that should be eliminated by the algorithm. Furthermore, in order to see how well our implementation will work with realistic data from the OQRNG generator, we will sample streams of bits directly from the source.

\subsubsection{5.1 Iterative approach}\label{iterative-approach}

Implementation of Toeplitz extraction will begin by a naive implementation not optimized for speed, but rather for accuracy. This implementation will be executed on a separate computer in order to produce the correct random number for every provided bitstring. These will be used as our baseline for accuracy for future iterations.

This naive implementation will then be flashed to our microcontrollers, beginning with Teensy 4.1 as this is the more capable of the microcontrollers used for this experiment. Code to measure the execution speed in microseconds will be implemented and tested before we load the naive implementation on said microcontroller. Each iteration will consist of incremental improvements to the algorithm. Our initial investigation has shown several avenues for improving the throughput of the algorithm, and each planned iteration will be discussed briefly. Depending on the empirical results of these iterations, there may be a need for further optimization iterations other than those listed.

\textbf{Iteration 1 - Naive implementation}: The naive implementation consists of a Toeplitz matrix \(T\) and key \(k\), acquired from fixed slices of the bitstream. Then, matrix-vector multiplication will be performed using \(T\) and \(k\), thus performing Toeplitz extraction -- which finally will eliminate jitter and produce a random number. The implementation of this in code will be quite simple, using nested loops to iterate over all values and multiply them. We expect that the throughput of this naive implementation will be far away from optimized iterations, and result in \(O(n^2)\) complexity.

\textbf{Iteration 2 - Efficient data structures}: The naive implementation will not utilize efficient data structures for maximum speed, rather it will likely use fixed-size arrays which need to be iterated over for a time complexity of \(O(n^2)\). Our hypothesis is that more efficient data structures may allow data to be processed in a far more efficient manner -- however, this scenario poses a risk for greater memory complexity which may not be suitable for use on microcontrollers. Which data structures might be most feasible will be evaluated during this iteration and as such is not preordained.

\textbf{Iteration 3 - Bitshifting}: We expect that Toeplitz extraction can be significantly improved with bitwise XOR operations, thus reducing overhead and memory usage. Instead of explicitly constructing the Toeplitz matrix, a right-shift operation can be used to dynamically reconstruct matrix rows. Additionally, matrix-vector multiplication can be improved using XOR operations to extract bits more efficient, thus minimizing unnecessary computations.

\textbf{Iteration 4 - Batching}: Finally, we consider the concept of batching larger amounts of bits for processing. Consider that an input buffer of bits is read from the ADC and stored, waiting for processing. Rather than taking 64 bits, and shifting them one by one, we can take two batches of 64 bits and multiply them directly using XOR bitshifting -- eliminating the need to process these bit-by-bit. As we will have a constant stream of bits from the ADC, we theorize that this method will allow for better performance in real-time processing over reading individual bits.

\textbf{Iteration 5 - ARM Hardware instructions}: Rather than performing bitshifting operations in the code itself, certain microcontrollers come equipped with a separate processor specifically for bitshift-operations. Offloading the shifting to these processors rather than running them on the main CPU may allow faster processing of the data than performing the shifting in the code itself. However, this operation isn't natively supported by Arduino, and will potentially lead to extreme rewrites of the code which may prove too time consuming to do for two microcontrollers (as the code will not be reusable between controllers). As such, this optimization may only be done for one controller or left for future work.

Each scenario will first be executed on a single thread, and multiple threads may theoretically be executed concurrently to further optimize the data. The feasibility of this highly depends on the performance of every individual implementation. Depending on the results during our experimentation, this may yield yet another iteration.

\subsubsection{5.2 Evaluation}\label{evaluation}

In order to evaluate the efficacy of each iteration, we will test the code by processing our pre-determined bitstreams, saving the processed bits in a binary file and evaluating this file with the command line utility \texttt{ent}\footnote{\href{https://manpages.ubuntu.com/manpages/trusty/man1/ent.1.html}{Manual page for \texttt{ent}, accessed 2025-04-23.}}, which is a simple way of evaluating the randomness of binary data.

As discussed in section 3, the hardware used will impose a clear bound on how quickly our implementation needs to process the bits in order to match the speed of the ADC and the output speed of the USB-port in \(Mbit/s\). In order to verify accuracy, the output of random numbers should be identical for each sample bitstring tested. Certain iterations might, however, use larger parts of the bitstring, and thus produce a slightly variable result.

@@

\[
T_{\mathrm{ms}}
= \frac{64}{24}\times10^{-3}\ \mathrm{ms}
\approx 2.667\times10^{-3}\ \mathrm{ms}
=2.667\ \mu\mathrm{s}.
\]

@@

\subsection{6 EXPERIMENTATION}\label{experimentation}

Each iteration results in an implementation written in C++, all of which will be supplied as an appendix to this paper. Furthermore, to simplify the implementation process, we will use the Arduino framework to communicate with the hardware. Whereas this introduces some overhead, it drastically reduces the time spent per iteration, and is deemed to be a suitable starting point.

Initially, we created a script to simulate a stream of bits from a file, and pre-computed the expected results using a trusted version of Toeplitz extraction. With the expected results saved to a file on the host computer, each iteration could then be deployed to the MCU, and tested with the same data as the baseline. As the algorithm is fully deterministic with this uniform data, the exact same results in the same order should be produced by all iterations. The simulation script then simply asserts that the MCU produces the same random string of bits from the resulting iterations as with the baseline.

As our system does not rely on interrupts in any capacity, measuring the execution speed of the implemented Toeplitz extraction function can be done using the built in \texttt{micros()}-function provided by Arduino. This provides us with an accurate measurement of the execution time per function call. The average execution time per test set over each iteration will be presented in \textbf{\texttt{TABLE\ 1}}, along with the fastest and slowest executions per iteration respectively. Furthermore, the entropy value on the processed bitstream will be noted in the table to demonstrate that the important measurement, e.g.~the randomness, still holds true between iterations.

For our tests, we used six files with varying degrees of patterns in the bitstring, ranging from quite disastrous to acceptable. We constructed a simple test script to facilitate easier testing of each iteration. This script allowed us to easily test each implementation by automating several tasks -- generating a baseline binary file with data processed by a good, proven implementation of Toeplitz extraction, giving us a way to ensure each iteration still produces the correct result. Furthermore, this script allows us to define different sample sizes for the MCU to utilize, and can extract the amount of time the algorithm takes to execute (\emph{see table 1}) per each sample size.

Below we'll detail the results for each iteration.

\newpage

\subsection{\texorpdfstring{RESULTS }{RESULTS }}\label{results}

All iterations was tested on 6 different input datasets, with some results varying less than 0.1 microseconds. These results with such slight variations have been omitted from the table. All the tables shown below has been tested on the dataset \texttt{biased\_bits\_bad.bin} which can be found as an appendix -- and whereas more files have been used, this file has been arbitrarily selected to present the data.

Due to compatibility issues, Pi Pico 2 was tested on different machine, creating further variation in the results -- about 1 microsecond.

\subsubsection{Iteration 1}\label{iteration-1}

The first iteration brought few surprises, as it ran quite well on Teensy. To our surprise, it executed well enough on Pi Pico 2 as well, albeit with \textasciitilde{}\(8\times\) slower execution speed. This first implementation is the same one used to generate the baseline results, and used to ensure following iterations all generate correct output.

Worth noting is that this iteration (\emph{and all following iterations that generate the same result}) have achieved entropy scores from \texttt{ent} that are greater than 7.99 for all tested data while ensuring it cannot be compressed further. This ensures that the algorithm has successfully removed as much of the patterns as can be reasonably expected from the raw binary data while maintaining the random nature of the data.

@@

\subsubsection{\texorpdfstring{Iteration 2 }{Iteration 2 }}\label{iteration-2}

Although the updated version included enhanced data structures, theoretical and practical analysis indicated that these modifications do not provide advantages for this project. Due to the small size of the input data (64, 512, 1024 bits) - using plain, contiguous memory layouts such as \texttt{std::vector\textless{}int\textgreater{}} proves to be very efficient in terms of access speed and memory usage. Current C++ compilers can effectively optimize these simple structures, often generating faster and more compact machine code than manually optimised alternatives. Introducing additional data structures or indirection would only add overhead without improving performance.

From an algorithmic perspective, Toeplitz extraction inherently requires a double summation loop, which leads to a time complexity of \(O(n^2)\) no matter the optimization applied internally. Thus, making attempts to change it would not result in lower asymptotic cost.

In summary, given the size and type of this system, a straightforward and minimal approach is not just adequate, but also recommended.

\subsubsection{Iteration 3}\label{iteration-3}

Coming soon!

>>>>

Further details about how this signal is produced and sampled will be introduced in section 2 and 3, and builds on the work of Clason \cite{Clason2023}. This optical signal needs to be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). Some processing of the data has to be done in order to ensure that the data is workable and to remove potential deterministic patterns from these random bits. One method for this post-processing we will explore in this work is Toeplitz extraction \cite{toeplitz}, typically performed on the host computer utilizing the randomly generated numbers. While this system could very well be implemented on physically larger hardware, thus avoiding the constraints that limited hardware introduces, Clason proposes a simpler and cheaper way to achieve QRNG \cite{Clason2023}. In keeping with this, utilizing microcontrollers rather than a host computer for processing the raw bits extracted further helps to keep the costs low and the solution reasonably complex. Additionally, this allows the system to be self-contained, for instance a simple and portable USB-device which both generates and processes true random numbers. Further details regarding the microcontrollers used in our work will be outlined in section 3.

However, due to the lower processing power of the average microcontroller, any implementation of Toeplitz extraction needs to work quickly and efficiently despite this hardware constraint. Experimenting with efficient implementations of this well-defined algorithm is the main focus of this thesis, and the key research areas explored in this work are as follows:

@@

\textbf{Research area 2 (\emph{RA2})}: Can we ensure that the output of random numbers is not primarily limited primarily by our implementation, but rather limited by the processing power or the USB transfer speed of the microcontroller, alternatively by the ADC?

@@

Section 2 of this article will introduce the theory that allows for QRNG, and how this will be utilized in our works. Section 3 delves further into the hardware and algorithms our work will use, with related works in optimizing Toeplitz extraction listed under section 4. Section 5 will present our methodology and implementation strategy, as well as some limitations imposed on our work. Finally, section 7 will present the results of our experimentation.

@@

Microcontrollers (MCUs) are compact and low-power computing devices designed for embedded systems and real-time operations, and suitable as a processing unit for the purposes of this work. Unlike general purpose CPUs, an MCU integrates a processor, memory and peripherals (\emph{such as an ADC, for instance}) into a single chip. Furthermore, modern MCUs often feature advanced microarchitectural elements to enhance processing capabilities on single threads (\emph{such as dual-issue superscalar architectures, allowing the MCU to run several instructions per CPU cycle}), making them suitable candidates for the post-processing required for OQRNG-data.

@@

In our work, we intend to use Teensy 4.1\footnote{\href{https://www.pjrc.com/store/teensy41.html}{Teensy developer documentation, accessed 2025-02-27}}, based on the ARM Cortex-M7. This MCU is especially suitable for computationally demanding tasks involving randomness extraction due to its dual-issue superscalar architecture and Digital Signal Processing (\emph{DSP}) capabilities. The floating-point unit\footnote{\href{https://www.sciencedirect.com/topics/computer-science/floating-point-unit}{ScienceDirect Journals \& Books, accessed 2025-03-13}} (FPU) and Single Instruction, Multiple Data (\emph{SIMD}) style DSP instructions improves how quickly it can perform bitwise and arithmetic tasks, which are crucial for quick Toeplitz extraction. SIMD-controlled DSP architectures, as described by Han et al. \cite{simd-dsp}, leverage parallel vectorized computation to accelerate matrix operations -- making them highly effective for Toeplitz matrix-vector multiplications.

In order to evaluate how efficent our implementation can become, our aim is to try our implementation on other MCUs with varying levels of power and hardware support. Whereas Teensy 4.1 is our primary development platform which we will evaluate closely, we aim to run our implementations on Raspberry Pi Pico 2\footnote{\href{https://datasheets.raspberrypi.com/pico/pico-2-product-brief.pdf}{Raspberry Pi Pico 2 documentation, accessed 2025-03-13}} as well as ESP32-S3\footnote{\href{https://www.espressif.com/sites/default/files/documentation/esp32-s3_datasheet_en.pdf}{ESP32-S3 documentation, accessed 2025-03-13}}. Due to the lower computational power of these MCUs, there may be significant issues in utilizing these weaker models, yet they are significantly cheaper and easier to access. Testing of these will consist solely of running the implementation on these controllers and measuring execution speed and correctness of the output.

\subsubsection{3.4 Toeplitz extraction}\label{toeplitz-extraction}

@@

Let X be the input array (length n)
Let T be the seed matrix (length n + m – 1)
Let Y be a new array of length m

@@

The main focus of this work is implementing this algorithm as efficiently as possible on our MCUs, and as such, several optimization efforts need to be taken into account during our experimentation.

@@

\subsection{4 RELATED WORKS}\label{related-works}

@@

However, in order to ensure the results work with varying data, we cannot limit ourselves to simply one stream of bits. The main point of the algorithm is to remove patterns in the bitstream that may lead to less randomized results. As such, we will sample several bitstreams from the OQRNG-device to use for our tests -- each with varying degrees of repeated patterns that should be eliminated by the algorithm. All bitstrings tested are available as an appendix to this paper.

\subsubsection{5.1 Implementation strategy}\label{implementation-strategy}

Our implementation of the Toeplitz extractor followed a structured, iterative approach divided into two phases. Phase one focused on exploring performance improvements through incremental algorithmic changes. Phase two then addressed architectural inefficiencies identified during phase one. Each phase then consisted of a number of individual iterations, in which performance was evaluated in terms of both correctness and execution time on the target hardware.

\paragraph{5.2 Phase one}\label{phase-one}

Initially, we require a ``naive'' version designed to prioritize correctness over speed. This version was first executed on a separate computer to generate reference output for various input bitstrings, which were later used as accuracy baselines. The naive implementation was then flashed onto the Teensy 4.1 microcontroller, where execution time was measured in microseconds. Each subsequent iteration introduced controlled modifications aimed at improving throughput.

The initial implementation followed the pseudocode described in section 3.4, using matrix multiplication over raw input and seed data. It relied on \texttt{std::vector\textless{}int\textgreater{}} for storage and used nested loops to compute each output bit. From this point, this implementation was improved over the coming iterations, and the new implementation verified in the same manner as the initial version.

\textbf{Iteration 1 - Data structures:} This iteration kept the same algorithmic logic as the initial implementation, but experimented with data structures such as raw pointers and hash maps to enhance the use of \texttt{std::vector}. The goal was to reduce or eliminate the reliance on nested loops.

\textbf{Iteration 2 - Bitshifting:} Basic bitwise operation was introduced to replace arithmetic whenever possible. Multiplication was replaced with logical \texttt{AND\ \&} and modulo operations with bit masking \texttt{\&\ 1}. The goal was to reduce the number of instructions and improve per-bit processing speed.

\textbf{Iteration 3 - Batching and Hardware optimization:} This iteration focused on optimizing performance through batching and the use of ARM-native instructions, beginning by testing batching alone. This was followed by isolated use of ARM instructions such as \texttt{\_\_builtin\_popcountll()} (\emph{which counts the number of set bits in an unsigned integer}). After establishing their individual effect we combined both techniques, multiple batch sizes were tested to determine their impact. More details and benchmarks for each configuration can be found in the section 6.

\paragraph{5.3 Phase two}\label{phase-two}

Phase two focused on addressing inefficiencies and design issues that were unintentionally introduced during earlier iterations. Rather than continuing with new algorithmic ideas, this phase aimed to identify and fix structural problems. Several assumptions from phase one were re-evaluated -- such as the benefits of certain data structures or abstractions.

\textbf{Iteration 4 - Loop unrolling:} This iteration focused on reducing the number of loops in the extractor by manually unrolling repeated operations. The goal was to decrease overhead created by loops. Whereas this operation is commonly done by compiler optimization, manually performing this guarantees that we unroll the loops rather than leaving it to the compiler.

\textbf{Iteration 5 - Removal of vector usage:} This iteration removed \texttt{std::vector} in favor of fixed-size types like \texttt{uint32\_t} and \texttt{uint64\_t} to reduce the overhead introduced by creating and populating this complex data structure.

\textbf{Iteration 6 - Data type exploration:} Following the removal of vectors, this iteration explored alternative static data types to determine the most efficient structure for storing input and seed data.

\textbf{Iteration 7 - Re-implementation of bitshifting:} In this final iteration, bit shifting techniques were reintroduced into the now simplified implementation. With high-level abstractions removed and memory structures streamlined, bit-level operations such as shifting and masking were applied.

\subsubsection{5.4 Evaluation}\label{evaluation}

To evaluate the correctness of each implementation, a baseline was generated as discussed in section 5.2. Using the naive, initial implementation to process bits and saving for later evaluation gave us a source of truth against which to compare following iterations. To verify that the algorithm successfully removed the patterns it should, we verified the measured entropy score with the command line utility \texttt{ent}\footnote{\href{https://manpages.ubuntu.com/manpages/trusty/man1/ent.1.html}{Manual page for \texttt{ent}, accessed 2025-04-23.}}.

We created a script to facilitate easier testing, which is attached to this paper. Using this script, all bitstrings used for testing can be evaluated against the baseline, ensuring that the output from the new iteration matches the baseline exactly. Furthermore, the script also provides the average execution time of only the Toeplitz extraction in microseconds.

@@

\[
T_{\mathrm{ms}}
= \frac{64}{24}\times10^{-3}\ \mathrm{ms}
\approx 2.667\times10^{-3}\ \mathrm{ms}
=2.667\ \mu\mathrm{s}.
\phantom{12}(3)
\]

In section 6, this calculation will be used to derive the execution speed of the various iterations.

@@

\subsection{6 RESULTS}\label{results}

@@

Table 1 presents the execution speeds of each iteration as listed in sections 5.3 and 5.4, alongside the exectution speed in Mbit/s as calculated by utilizing the average value in (3). My thought then is that we here talk about the results and what they demonstrate, before moving on to draw conclusions and attempt to answer our research questions in section 7.
